---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SteadyStateBVAR

<!-- badges: start -->
<!-- badges: end -->

This package provides estimation and forecasting of the Steady-State BVAR(p) model by Mattias Villani.

## Installation

You can install the development version of SteadyStateBVAR from [GitHub](https://github.com/) with:

``` r
#install.packages("devtools")
devtools::install_github("markjwbecker/SteadyStateBVAR")
```

## Example

Let us first load the library and some Swedish macro data.

```{r example, message=FALSE, warning=FALSE}
library(SteadyStateBVAR)
data("swe_macro")
```

Lets plot the Swedish macro data consisting of CPIF inflation, unemployment rate and the 3-month interest rate.

```{r cars}
plot.ts(Y)
```
We now do some setup

```{r setup}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
