---
output:
  github_document:
    toc: true
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SteadyStateBVAR

<!-- badges: start -->
<!-- badges: end -->

With this package the user can estimate the Steady-State BVAR(p) model by Mattias Villani.

## Installation

You can install the development version of SteadyStateBVAR with:

```{r, eval=FALSE}
remotes::install_github(
  "markjwbecker/SteadyStateBVAR",
  ref = "dev",
  force = TRUE,
  upgrade = "never"
)

```
## Introduction

The Steady State BVAR($p$) model (Villani, 2009) is

$$
y_t = \Psi x_t + A_1(y_{t-1}-\Psi x_{t-1})+\dots+A_p(y_{t-p}-\Psi x_{t-p})+u_t
$$

where $y_t$ is a $k$-dimensional vector of (endogenous) time series at time $t$, and $x_t$ is a $q$-dimensional vector of deterministic/exogenous variables at time $t$, and $u_t \sim N_k(0,\Sigma_u)$ with independence between time periods. Also, $A_\ell$ for $\ell=1,\dots,p$ is $(k \times k)$, and $\Psi$ is $(k \times q)$. Note here that


$$
E(y_t)=\mu_t=\Psi x_t
$$

is the **steady state**. Note that the current version of this package only allows for $x_t$ to either just contain a constant, or a constant and a dummy variable. We can stack the $A$ matrices in the $(kp \times k)$ matrix $\beta$

$$
\beta=
\begin{bmatrix}
A'_1 \\ 
\vdots  \\
A'_p
\end{bmatrix}
$$

We can then rewrite the model as a nonlinear regression (Karlsson, 2013)

$$
y_t' =x_t'\Psi' + \left[w_t'-q_t'(I_p \otimes \Psi') \right]\beta +u_t'
$$

where $w_t'=(y_{t-1}',\dots,y_{t-p}')$ is a $kp$-dimensional vector of lagged endogenous variables, and $q_t'=(x_{t-1}',\dots,x_{t-p}')$ is a $qp$-dimensional vector of lagged exogenous (deterministic) variables, $I_p$ is the $(p \times p)$ identity matrix and $\otimes$ denotes the Kronecker product. This is how the likelihood is written in the Stan code. The goal is to estimate $\beta, \Psi$ and $\Sigma_u$, therefore priors are needed. Starting with $\beta$, we use the Minnesota prior where


$$
\textrm{vec}(\beta) \sim N_{kpk}\left[\theta_\beta,\Omega_\beta\right]
$$

First for $\theta_\beta$, the Minnesota prior sets all prior means for the elements in $\beta$ to $0$, except for the elements that relate to the first own lags of the variables, which are often set to $0.9$ or $1$ for level variables or also to $0$ for growth rate variables. Further, $\Omega_\beta$ is a diagonal matrix containing the prior variances for the elements in $\beta$. The prior is constructed such that for the autoregressive coefficient $A_{\ell}^{(i,j)}$ which is element $\left(i,j\right)$ of $A_{\ell}$ for $\ell=1,\dots,p,$ the prior variance is given by


$$
\textrm{Var}\left(A_{\ell}^{(i,j)}\right)=
\begin{cases}
\left(\frac{\lambda_1}{\ell^{\lambda_3}}\right)^2 & \text{if } i = j \\
\left(\frac{\lambda_1 \lambda_2}{\ell^{\lambda_3}}\right)^2 \frac{\sigma_i^2}{\sigma_j^2}& \text{if } i \neq j
\end{cases}
$$


Here $\lambda_1$, $\lambda_2$ and $\lambda_3$ are scalar hyperparameters known as the overall tightness, the cross-equation tightness and the lag decay rate. Furthermore, $\sigma_i^2$ is the $(i,i)$:th element of $\Sigma_u$, which we do not know, and therefore replace with an estimate. In this package, it is replaced by the least squares residual variance from a univariate autoregression for variable $i$ with $p$ lags (including the constant and dummy variable if applicable). Moving on to $\Psi$ the prior we use is


$$
\textrm{vec}(\Psi) \sim N_{kq}\left[\theta_\Psi,\Omega_\Psi\right]
$$

This is really the core of the Steady State BVAR model. In $\theta_\Psi$ we specify our prior beliefs of the location of the steady state, and in $\Omega_\Psi$, which we assume to be a diagonal matrix, we specify our degree of certainty in those prior beliefs. At last, the prior for $\Sigma_u$ is


$$
\Sigma_u \sim IW(V_0,m_0)
$$

Here $V_0$ is the scale matrix and $m_0\geq k+2$ are the degrees of freedom. We will specify an uninformative prior by setting $V_0=(m_0-k-1)\hat{\Sigma}_u$ where $\hat{\Sigma}_u$ is the least squares estimate from the VAR($p$) (including the constant and dummy variable if applicable), and $m_0=k+2$. However if the user wants, the usual noninformative Jeffreys prior

$$
p(\Sigma_u) \propto\left|\Sigma_u \right|^{-(k+1)/2}
$$

can be used instead for $\Sigma_u$.

## Example 1 (Villani, 2009)

We will now replicate the model in the empirical analysis in Section 4.1 in Villani (2009). First let us load the library and also load the data

```{r example, message=FALSE, warning=FALSE}
library(SteadyStateBVAR)
data("villani2009")
yt <- villani2009
```
The data set contains quarterly data for Sweden over the time period 1980Q1–2005Q4. The seven variables are: trade-weighted measures of foreign GDP growth $(\Delta y_f)$, CPI inflation $(\pi_f)$ and the 3-month interest rate $(i_f)$, the corresponding domestic variables ($\Delta y$, $\pi$ and $i$), and the level of the real exchange rate defined as $q=s+p_f-p$, where $p_f$ and $p$ are the foreign and domestic CPI levels (in logs) and $s$ is the (log of the) trade weighted nominal exchange rate. As such we have

$$
y_t=
\begin{pmatrix}
\Delta y_f \\
\pi_f \\
i_f \\
\Delta y \\
\pi \\
i \\
q
\end{pmatrix}
$$

Also, we will leave out the last two observations, so the user can compare the forecasts produced here to the last forecasts seen in Figure 1-3 in Villani (2009), to verify that this implementation works correctly.

```{r, fig.width=8, fig.height=8}
yt <- ts(yt[1:102, ], start = start(yt), frequency = frequency(yt))
plot.ts(yt)
```

Also, let us create the bvar object which we will use throughout here.

```{r, fig.width=8, fig.height=8}
bvar_obj <- bvar(data = yt)
```

To model the Swedish financial crisis at the beginning of the 90s and the subsequent shift in monetary policy to inflation targeting and flexible exchange rate, $x_t$ (deterministic variables at time $t$) includes a constant term and a dummy for the pre-crisis period, i.e.


$$
x_{t}' =
\begin{cases}
\begin{pmatrix}1 & 1\end{pmatrix} & \text{if } t \le 1992Q4 \\
\begin{pmatrix}1 & 0\end{pmatrix} & \text{if } t > 1992Q4
\end{cases}
$$

```{r, fig.width=8, fig.height=8}
bp = 52 #breakpoint at 1992Q4
dum_var <- c(rep(1,bp), rep(0,nrow(yt)-bp)) #1 if t<=1992Q4, 0 if t>1992Q4
```

To formulate a prior on $\Psi$, note that the specification of $x_t$ implies the following parametrization of the steady state:


$$
\mu_t =
\begin{cases}
\psi_1 + \psi_2 & \text{if } t \le 1992Q4 \\
\psi_1 & \text{if } t > 1992Q4
\end{cases}
$$

where $\psi_i$ is the $i$:th column of $\Psi$. Now to setup the model. Although it is not mentioned which lag length is used in Villani (2009) we assume it is $p=4$.

```{r, fig.width=8, fig.height=8}
bvar_obj <- setup(bvar_obj,
                  p=4,
                  deterministic = "constant_and_dummy",
                  dummy = dum_var)
```

Now let us specify the priors. We first consider $\beta$. We choose the same values for the hyperparameters as in Villani (2009), i.e. overall tightness $\lambda_1=0.2$, cross equation tightness $\lambda_2=0.5$ and lag decay rate $\lambda_3=1$. We then specify the prior means for the first own lags of the variables. For variables in growth rates, we set the prior mean to $0$, for variables in levels, we set the prior mean to 0.9.

```{r, fig.width=8, fig.height=8}
lambda_1 <- 0.2
lambda_2 <- 0.5
lambda_3 <- 1.0

fol_pm=c(0,   #delta y_f
         0,   #pi_f
         0.9, #i_f
         0,   #delta y
         0,   #pi
         0.9, #i
         0.9  #q
         )
```

Now moving on to $\Psi$ for the steady state priors, we set them according to the 95% prior probability intervals in Table I in Villani (2009). We first note that for our data here, the growth rate variables ($\Delta y_f, \pi_f, \Delta y, \pi$) are specified in terms of quarterly rates of change, i.e. for a variable $z$, the quarterly growth rate is $100 \left(\ln z_t - \ln z_{t-1}\right)$. The 95% prior probability intervals in Table I are specified in terms of annualized quarterly growth rates $400 \left(\ln z_t - \ln z_{t-1}\right)$.

The 'ppi()' function is useful here. Simply input the desired 95\% prior probability interval on the annualized scale with 'annualized_growthrate=TRUE' and we get the corresponding prior mean and variance in terms of quarterly rate of change. Of course we could also just annualize our data before, and set 'annualized_growthrate='FALSE'. So now we do this for all steady state coefficients. Again, see Table I in Villani (2009) for the 95% prior probability intervals.

```{r, fig.width=8, fig.height=8}
#psi_1 = Psi col 1
#psi_2 = Psi col 2

theta_Psi <- 
  c(
  ppi( 2.00,  3.00,  annualized_growthrate=TRUE)$mean,   #psi_1: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$mean,   #psi_1: pi_f
  ppi( 4.50,  5.50,  annualized_growthrate=FALSE)$mean,  #psi_1: i_f
  ppi( 2.00,  2.50,  annualized_growthrate=TRUE)$mean,   #psi_1: y
  ppi( 1.70,  2.30,  annualized_growthrate=TRUE)$mean,   #psi_1: pi
  ppi( 4.00,  4.50,  annualized_growthrate=FALSE)$mean,  #psi_1: i
  ppi( 3.85,  4.00,  annualized_growthrate=FALSE)$mean,  #psi_1: q
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$mean,   #psi_2: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$mean,   #psi_2: pi_f
  ppi( 1.50,  2.50,  annualized_growthrate=FALSE)$mean,  #psi_2: i_f
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$mean,   #psi_2: y
  ppi( 4.30,  5.70,  annualized_growthrate=TRUE)$mean,   #psi_2: pi
  ppi( 3.00,  5.50,  annualized_growthrate=FALSE)$mean,  #psi_2: i
  ppi(-0.50,  0.50,  annualized_growthrate=FALSE)$mean   #psi_2: q
  )

Omega_Psi <- 
  diag(
  c(
  ppi( 2.00,  3.00,  annualized_growthrate=TRUE)$var,    #psi_1: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$var,    #psi_1: pi_f
  ppi( 4.50,  5.50,  annualized_growthrate=FALSE)$var,   #psi_1: i_f
  ppi( 2.00,  2.50,  annualized_growthrate=TRUE)$var,    #psi_1: y
  ppi( 1.70,  2.30,  annualized_growthrate=TRUE)$var,    #psi_1: pi
  ppi( 4.00,  4.50,  annualized_growthrate=FALSE)$var,   #psi_1: i
  ppi( 3.85,  4.00,  annualized_growthrate=FALSE)$var,   #psi_1: q
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$var,    #psi_2: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$var,    #psi_2: pi_f
  ppi( 1.50,  2.50,  annualized_growthrate=FALSE)$var,   #psi_2: i_f
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$var,    #psi_2: y
  ppi( 4.30,  5.70,  annualized_growthrate=TRUE)$var,    #psi_2: pi
  ppi( 3.00,  5.50,  annualized_growthrate=FALSE)$var,   #psi_2: i
  ppi(-0.50,  0.50,  annualized_growthrate=FALSE)$var    #psi_2: q
  )
  )
```

Finally for $\Sigma_u$ we will use the noninformative Jeffreys prior $\left|\Sigma_u \right|^{-(k+1)/2}$, as done in Villani (2009). We just input everything to the priors function. Note here that the function automatically creates $\theta_\beta$ and $\Omega_\beta$.


```{r, fig.width=8, fig.height=8}
bvar_obj <- priors(bvar_obj,
                   lambda_1,
                   lambda_2,
                   lambda_3,
                   fol_pm,
                   theta_Psi, 
                   Omega_Psi,
                   Jeffrey=FALSE)
```

Like in Villani (2009), to incorporate that Sweden is a small economy and therefore not likely to affect the foreign economy, we restrict the upper right submatrix in each $A_\ell$ for $\ell =1,\dots,k$ or equivalently restricting the bottom left $A_\ell'$ for $\ell =1,\dots,k$, to the zero matrix.

```{r, fig.width=8, fig.height=8}
p <- bvar_obj$setup$p
k <- bvar_obj$setup$k
kf <- 3 #first 3 variables are foreign in yt

restriction_matrix <- matrix(1, k*p, k)

for(i in 1:p){
  rows <- ((i-1)*k + kf + 1) : (i*k)
  cols <- 1:kf
  restriction_matrix[rows, cols] <- 0
}
restriction_matrix
```
We can look at the restriction matrix for $\beta$ to see which elements we restrict to zero. Since the priors means for these elements are zero, we do the restriction by setting the relevant prior variances in $\Omega_\beta$ to be very small. We just input our $(kp \times k)$ restriction matrix to the 'restrict' function like below:

```{r, fig.width=8, fig.height=8}
bvar_obj <- restrict(bvar_obj, restriction_matrix)
```

At last we need to supply our forecast horizon $H$, and also the deterministic variables for the future periods and then we fit the model.

```{r, fig.width=8, fig.height=8, results='hide'}
bvar_obj$predict$H <- 12
bvar_obj$predict$X_pred <- cbind(rep(1, 12), 0)
```

Then we can fit the model.

```{r, fig.width=8, fig.height=8, results='hide'}
bvar_obj <- fit(bvar_obj,
                iter = 5000,
                warmup = 2500,
                chains = 2)
```

Let us look at the posterior mean of $\beta$, $\Psi$ and $\Sigma_u$

```{r, fig.width=8, fig.height=8}
summary(bvar_obj)
```

Note that 'bvar_obj\$fit\$stan' is an object of class stanfit

```{r, fig.width=8, fig.height=8}
bvar_obj$fit$stan
```

Lets plot the forecasts. Lets select a 95% prediction interval and the mean of the posterior as the point forecast. For variables in quarter on quarter growth rates, we transform the historical data and predictions to yearly growth rates with 'growth_rate_idx' where we specify the index of the growth rate variables in $y_t$.

```{r, fig.width=8, fig.height=8}
bvar_obj <- forecast(bvar_obj,
                     ci = 0.95,
                     fcst_type = "mean",
                     growth_rate_idx = c(4,5),
                     plot_idx = c(4,5,6))
```

We can also do some impulse response analysis. We can choose between the orthogonalized impulse response function (OIRF) and the generalized impulse response function (GIRF).

```{r, fig.width=8, fig.height=8}
par(mfrow=c(2,2))
irf <- IRF(bvar_obj,
           lag=20,
           response=4,
           shock=6,
           method="OIRF",
           ci=0.68)

irf <- IRF(bvar_obj,
           lag=20,
           response=4,
           shock=6,
           method="GIRF",
           ci=0.68)

irf <- IRF(bvar_obj,
           lag=20,
           response=5,
           shock=6,
           method="OIRF",
           ci=0.68)

irf <- IRF(bvar_obj,
           lag=20,
           response=5,
           shock=6,
           method="GIRF",
           ci=0.68)
```

If desired, the user can estimate the model with a Gibbs sampler instead and then repeat the analysis above.

```{r, fig.width=8, fig.height=8}
# bvar_obj <- fit_gibbs(bvar_obj,
#                       iter = 10000,
#                       warmup = 5000)

# summary(bvar_obj)

# bvar_obj <- forecast(bvar_obj,
#               ci = 0.95,
#               fcst_type = "mean",
#               growth_rate_idx = c(1,2,4,5),
#               plot_idx = c(4,5,6),
#               estimation="gibbs")

# irf <- IRF(bvar_obj,
#            lag=20,
#            response=4,
#            shock=6,
#            method="OIRF",
#            ci=0.68,
#            estimation="gibbs")
```

It will give very similar results (as it should be). We will test if they give similar results in the following section.

## Example 2 (Gustafsson and Villani, 2025)

Now I will estimate the same model in Section 3.3 in Gustafsson and Villani (2025), and "replicate" Figure 10-11. But now instead of blue = Variational inference, I will do blue = Stan (NUTS sampler), and red = Gibbs sampling. The data used in Gustafsson and Villani (2025) is the same as in Gustafsson, Villani and Stockhammar (2023), from which I obtained the data.

```{r, fig.width=8, fig.height=8}
# rm(list = ls())
# data("GustafssonVillaniStockhammar2023")
# yt <- GustafssonVillaniStockhammar2023
# plot.ts(yt)
```

Create the object

```{r, fig.width=8, fig.height=8}
# bvar_obj <- bvar(data = yt)
```

Only a constant this time, thus $q=1$. The lag order $p$ is not specified, but looking at Gustafsson, Villani and Stockhammar (2023) we can sort of infer that they use $p=2$.

```{r, fig.width=8, fig.height=8}
# bvar_obj <- setup(bvar_obj,
#                   p=2,
#                   deterministic = "constant",
#                   dummy = NULL)
```

The hyperparameters are from Gustafsson, Villani and Stockhammar (2023), and are $\lambda_1=0.27$, $\lambda_2=0.43$ and $\lambda_3=0.76$. The prior means on the first own lag of the GDP-deflator and Fed funds rate are set to 0.6.

```{r, fig.width=8, fig.height=8}
# lambda_1 <- 0.27
# lambda_2 <- 0.43
# lambda_3 <- 0.76
# 
# fol_pm=c(0,   #GDPC1
#          0.6, #GDPCTPI
#          0.6, #FEDFUNDS
#          0,   #PCECC96
#          0,   #GPDIC1
#          0,   #HOANBS
#          0    #AHETPIx
#          )
```

For the steady state coefficients, the prior probability intervals (normal distribution) are mean plus/minus one standard deviation (i.e. a roughly 68% prior probability interval)

```{r, fig.width=8, fig.height=8}
# 1-2*(1-pnorm(1))
```

Note that we only have a constant now, so then $q=1$ and $\Psi$ only has one column $\psi_1$. Also the growth rate variables are already in the form $400 \left(\ln z_t - \ln z_{t-1}\right)$, so no need to annualize.

```{r, fig.width=8, fig.height=8}
# theta_Psi <- c(
#   ppi(2.5, 3.5, interval=0.6826895)$mean,   #psi_1: GDPC1
#   ppi(1.5, 2.5, interval=0.6826895)$mean,   #psi_1: GDPCTPI
#   ppi(4.3, 5.7, interval=0.6826895)$mean,   #psi_1: FEDFUNDS
#   ppi(2.3, 3.7, interval=0.6826895)$mean,   #psi_1: PCECC96
#   ppi(1.5, 4.5, interval=0.6826895)$mean,   #psi_1: GPDIC1
#   ppi(2.5, 3.5, interval=0.6826895)$mean,   #psi_1: HOANBS
#   ppi(1.5, 2.5, interval=0.6826895)$mean    #psi_1: AHETPIx
# )
# 
# Omega_Psi <- diag(
#   c(
#   ppi(2.5, 3.5, interval=0.6826895)$var,    #psi_1: GDPC1
#   ppi(1.5, 2.5, interval=0.6826895)$var,    #psi_1: GDPCTPI
#   ppi(4.3, 5.7, interval=0.6826895)$var,    #psi_1: FEDFUNDS
#   ppi(2.3, 3.7, interval=0.6826895)$var,    #psi_1: PCECC96
#   ppi(1.5, 4.5, interval=0.6826895)$var,    #psi_1: GPDIC1
#   ppi(2.5, 3.5, interval=0.6826895)$var,    #psi_1: HOANBS
#   ppi(1.5, 2.5, interval=0.6826895)$var     #psi_1: AHETPIx
#   )
# )
```

Lets put everything into the 'priors function'

```{r, fig.width=8, fig.height=8}
# bvar_obj <- priors(bvar_obj,
#                    lambda_1,
#                    lambda_2,
#                    lambda_3,
#                    fol_pm,
#                    theta_Psi,
#                    Omega_Psi,
#                    Jeffrey=TRUE)
```

We supply our forecast horizon and then estimate the model. We do it both with Stan and with the Gibbs sampler.

```{r, fig.width=8, fig.height=8,results='hide'}
# bvar_obj$forecasts$H <- 12
# bvar_obj$forecasts$X_pred <- matrix(rep(1, 12))
# 
# bvar_obj <- fit_gibbs(bvar_obj,
#                      iter=10000,
#                      warmup=5000)
# 
# bvar_obj <- fit_stan(bvar_obj,
#                      iter=5000,
#                      warmup=2500,
#                      chains=4)
```
Lets check the estimation results

```{r, fig.width=8, fig.height=8}
# summary(bvar_obj)
```

Now lets do Figure 10

```{r, fig.width=8, fig.height=8}
# Psi_draws_gibbs  <- bvar_obj$fit$gibbs$Psi_draws
# Psi_draws_stan <- rstan::extract(bvar_obj$fit$stan)$Psi
# 
# par(mfrow=c(2,2))
# 
# # Real GDP
# dens1 <- density(Psi_draws_gibbs[1,,])
# dens2 <- density(Psi_draws_stan[,1,])
# plot(dens1, xlab="Real GDP", main="", col="red", lwd=2, ylim=c(0,max(dens1$y,dens2$y)))
# lines(dens2, col="blue", lwd=2)
# legend("topright", legend=c("Gibbs", "Stan"), col=c("red", "blue"), lwd=2, bty="n")
# 
# # GDP deflator
# dens1 <- density(Psi_draws_gibbs[2,,])
# dens2 <- density(Psi_draws_stan[,2,])
# plot(dens1, xlab="GDP deflator", main="", col="red", lwd=2, ylim=c(0,max(dens1$y,dens2$y)))
# lines(dens2, col="blue", lwd=2)
# legend("topright", legend=c("Gibbs", "Stan"), col=c("red", "blue"), lwd=2, bty="n")
# 
# # Fed funds rate
# dens1 <- density(Psi_draws_gibbs[3,,])
# dens2 <- density(Psi_draws_stan[,3,])
# plot(dens1, xlab="Fed funds rate", main="", col="red", lwd=2, ylim=c(0,max(dens1$y,dens2$y)))
# lines(dens2, col="blue", lwd=2)
# legend("topright", legend=c("Gibbs", "Stan"), col=c("red", "blue"), lwd=2, bty="n")
# 
# # Real consumption
# dens1 <- density(Psi_draws_gibbs[4,,])
# dens2 <- density(Psi_draws_stan[,4,])
# plot(dens1, xlab="Real consumption", main="", col="red", lwd=2, ylim=c(0,max(dens1$y,dens2$y)))
# lines(dens2, col="blue", lwd=2)
# legend("topright", legend=c("Gibbs", "Stan"), col=c("red", "blue"), lwd=2, bty="n")
```

And Figure 11 (mean +/- 1 std deviation credible bands)

```{r, fig.width=8, fig.height=8}
# par(mfrow=c(2,2))
# forecast2(bvar_obj, fcst_type="mean", plot_idx=c(1), xlim=c(39.25,58), ylim=c(-8.5,7))
# forecast2(bvar_obj, fcst_type="mean", plot_idx=c(2), xlim=c(39.25,58), ylim=c(-1,4))
# forecast2(bvar_obj, fcst_type="mean", plot_idx=c(3), xlim=c(39.25,58), ylim=c(-1,6.5))
# forecast2(bvar_obj, fcst_type="mean", plot_idx=c(4), xlim=c(39.25,58), ylim=c(-3.5,6.5))
```


## References

Gustafsson, O., Villani, M., and Stockhammar, Pär. (2023). Bayesian optimization of hyperparameters from noisy marginal likelihood estimates. *Journal of Applied Econometrics*. 38(4), pp. 577–595.

Gustafsson, O., and Villani, M. (2025). Variational inference for steady-state BVARs. *arXiv preprint 	arXiv:2506.09271*.

Karlsson, S. (2013). Forecasting with Bayesian Vector Autoregression. In: Elliott, G. and Timmerman, A. (eds) *Handbook of Economic Forecasting*. Elsevier B.V. Vol 2, Part B., pp. 791-897.

Villani, M. (2009). Steady-state priors for vector autoregressions. *Journal of Applied Econometrics*. 24(4), pp. 630-650. 