---
output:
  github_document:
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SteadyStateBVAR

<!-- badges: start -->
<!-- badges: end -->

With this package the user can estimate the Steady-State BVAR(p) model by Mattias Villani.

## Installation

You can install the development version of SteadyStateBVAR with:

```{r, results='hide'}
#remotes::install_github("markjwbecker/SteadyStateBVAR", force = TRUE, upgrade = "never")
```
## Introduction

The Steady State BVAR($p$) model is

$$
y_t = \Psi x_t + A_1(y_{t-1}-\Psi x_{t-1})+\dots+A_p(y_{t-p}-\Psi x_{t-p})+u_t
$$

where $y_t$ is a $k$-dimensional vector of time series at time $t$, and $x_t$ is a $q$-dimensional vector of deterministic variables at time $t$, and $u_t \sim N_k(0,\Sigma_u)$ with independence between time periods. Also, $A_i$ for $i=1,\dots,p$ is $(k \times k)$, and $\Psi$ is $(k \times q)$. Note here that


$$
E(y_t)=\mu_t=\Psi x_t
$$

is the **steady state**. We can stack the $A$'s in the $(kp \times k)$ matrix $\beta$

$$
\beta=
\begin{bmatrix}
A'_1 \\ 
\vdots  \\
A'_p
\end{bmatrix}
$$

We can then rewrite the model as a nonlinear regression

$$
y_t' =x_t'\Psi' + \left[w_t'-q_t'(I_p \otimes \Psi') \right]\beta +u_t'
$$

where $w_t'=(y_{t-1}',\dots,y_{t-p}')$ is a $kp$-dimensional vector of lagged endogenous variables, and $q_t'=(x_{t-1}',\dots,x_{t-p}')$ is a $qp$-dimensional vector of lagged exogenous (deterministic) variables, $I_p$ is the $(p \times p)$ identity matrix and $\otimes$ denotes the Kronecker product. This is how the likelihood is written in the Stan code. The goal is to estimate $\beta, \Psi$ and $\Sigma_u$, therefore priors are needed. Starting with $\beta$, we use the Minnesota prior where


$$
\textrm{vec}(\beta) \sim N_{kpk}\left[\textrm{vec}(\beta_0),\Sigma_{\textrm{vec}(\beta)}\right]
$$

First for $\beta_0$, the Minnesota prior sets all elements to $0$, except for the elements that relate to the first own lags of the variables, which are often set to $0.9$ or $1$ for level variables or also to $0$ for growth rate variables. Further, $\Sigma_{\textrm{vec}(\beta)}$ is a diagonal matrix containing the prior variances for the elements in $\beta$. The prior is constructed such that for the autoregressive coefficient $A_{\ell}^{(i,j)}$ which is element $\left(i,j\right)$ of $A_{\ell}$ for $\ell=1,\dots,p,$ the prior variance is given by


$$
\textrm{Var}\left(A_{l}^{(i,j)}\right)=
\begin{cases}
\left(\frac{\lambda_1}{\ell}\right)^2 & \text{if } i = j \\
\left(\frac{\lambda_1 \lambda_2}{\ell}\right)^2 \frac{\sigma_i^2}{\sigma_j^2}& \text{if } i \neq j
\end{cases}
$$


Here $\lambda_1$ and $\lambda_2$ are scalar hyperparameters where the former is known as the overall tightness, and the latter as the cross-equation tightness. Furthermore, $\sigma_i^2$ is the $(i,i)$:th element of $\Sigma_u$, which we do not know, and therefore replace with an estimate. In this package, it is replaced by the least squares residual variance from a univariate autoregression for variable $i$ with $p$ lags (including any potential deterministic variables). Moving on to $\Psi$ the prior we use is


$$
\textrm{vec}(\Psi) \sim N_{kq}\left[\textrm{vec}(\Psi_0),\Sigma_{\textrm{vec}(\Psi)}\right]
$$

It is assumed that $\Sigma_{\textrm{vec}(\Psi)}$ is a diagonal matrix. At last, the prior for $\Sigma_u$ is


$$
\Sigma_u \sim IW(V_0,m_0)
$$

Here $V_0$ is the scale matrix and $m_0\geq k+2$ are the degrees of freedom. Since Stan does not "allow" the usual noninformative prior $\left|\Sigma_u \right|^{-(k+1)/2}$, we will instead specify an uninformative prior by setting $V_0=(m_0-k-1)\hat{\Sigma}_u$ where $\hat{\Sigma}_u$ is the least squares estimate from the VAR($p$) (including any potential deterministic regressors), and $m_0=k+2$.

## Example

We will now replicate the model in the empirical analysis in Section 4.1 in Villani (2009). First let us load the library and also load the data

```{r example, message=FALSE, warning=FALSE}
#remotes::install_github("markjwbecker/SteadyStateBVAR", force = TRUE, upgrade = "never")
library(SteadyStateBVAR)
data("villani2009")
```
The data set contains quarterly data for Sweden over the time period 1980Q1–2005Q4. The seven variables are: trade-weighted measures of foreign GDP growth $(\Delta y_f)$, CPI inflation $(\pi_f)$ and the 3-month interest rate $(i_f)$, the corresponding domestic variables ($\Delta y$, $\pi$ and $i$), and the level of the real exchange rate defined as $q=s+p_f-p$, where $p_f$ and $p$ are the foreign and domestic CPI levels (in logs) and $s$ is the (log of the) trade weighted nominal exchange rate. As such we have

$$
y_t=
\begin{pmatrix}
\Delta y_f \\
\pi_f \\
i_f \\
\Delta y \\
\pi \\
i \\
q
\end{pmatrix}
$$

The growth rate variables are specified in terms of quarterly rates of change, i.e. for a variable $z$, we have $100 \ln \left(\frac{z_t}{z_{t-1}}\right)$. To simplify the prior selection for the steady states later, we can multiply the growth rate series by $4$ to instead get annualized quarterly rates of change. We do that and then we can plot the data.

```{r, fig.width=8, fig.height=8}
growth_idx <- c(1, 2, 4, 5)
yt[, growth_idx] <- yt[, growth_idx] * 4
plot.ts(yt)
```

To model the Swedish financial crisis at the beginning of the 90s and the subsequent shift in monetary policy to inflation targeting and flexible exchange rate, $x_t$ (deterministic variables at time $t$) includes a constant term and a dummy for the pre-crisis period, i.e.


$$
x_{t}' =
\begin{cases}
\begin{pmatrix}1 & 1\end{pmatrix} & \text{if } t \le 1992Q4 \\
\begin{pmatrix}1 & 0\end{pmatrix} & \text{if } t > 1992Q4
\end{cases}
$$

```{r, fig.width=8, fig.height=8}
bp = 52 #breakpoint at 1992Q4
dum_var <- c(rep(1,bp), rep(0,nrow(yt)-bp)) #1 if t<=1992Q4, 0 if t>1992Q4
```

To formulate a prior on $\Psi$, note that the specification of $x_t$ implies the following parametrization of the steady state:


$$
\mu_t =
\begin{cases}
\psi_1 + \psi_2 & \text{if } t \le 1992Q4 \\
\psi_1 & \text{if } t > 1992Q4
\end{cases}
$$

where $\psi_i$ is the $i$:th column of $\Psi$. More on this soon. First we assume $p=4$ is a good lag length for our model. Then we do some setup.

```{r, fig.width=8, fig.height=8}
p=4
stan_data <- BVAR_setup(yt, p, deterministic="constant_and_dummy", dummy=dum_var)
```

Now let us specify the priors. First we choose overall tightness $\lambda_1=0.2$ and cross equation tightness $\lambda_2=0.5$. We then specify the prior means for the first own lags of the variables. For variables in growth rates, we set the prior mean to $0$, for variables in levels, we set the prior mean to 0.9.

```{r, fig.width=8, fig.height=8}
lambda1=0.2 #overall tightness
lambda2=0.5 #cross equation tightness

#first own lag prior mean
fol_pm=c(0,   #delta y_f
         0,   #pi_f
         0.9, #i_f
         0,   #delta y
         0,   #pi
         0.9, #i
         0.9  #q
)
```

Now for the steady state priors, we set them according to the 95% prior probability intervals in Villani (2009). The 'ppi()' function is useful here, simply input the 95% prior probability interval and then the function outputs the prior mean and variance.

```{r, fig.width=8, fig.height=8}
Psi_0 <- matrix(c(
                  ppi(2,3)$mean,
                  ppi(1.5,2.5)$mean,
                  ppi(4.5,5.5)$mean, 
                  ppi(2,2.5)$mean,
                  ppi(1.7,2.3)$mean,
                  ppi(4,4.5)$mean,
                  ppi(3.85,4)$mean,
                  ppi(-1,1)$mean,
                  ppi(1.5,2.5)$mean,
                  ppi(1.5,2.5)$mean,
                  ppi(-1,1)$mean,
                  ppi(4.3,5.7)$mean,
                  ppi(3,5.5)$mean,
                  ppi(-0.5,0.5)$mean
                  ), nrow=stan_data$k, ncol=stan_data$q)

vec_Psi_vars <- c(ppi(2,3)$var,
                  ppi(1.5,2.5)$var,
                  ppi(4.5,5.5)$var, 
                  ppi(2,2.5)$var,
                  ppi(1.7,2.3)$var,
                  ppi(4,4.5)$var,
                  ppi(3.85,4)$var,
                  ppi(-1,1)$var,
                  ppi(1.5,2.5)$var,
                  ppi(1.5,2.5)$var,
                  ppi(-1,1)$var,
                  ppi(4.3,5.7)$var,
                  ppi(3,5.5)$var,
                  ppi(-0.5,0.5)$var)
```
Let us check out the prior means in
```{r, fig.width=8, fig.height=8}
Psi_0
```
So if we are "after" 1992Q4, the steady state prior for variable $i$ is row $i$ column $1$. If we are before or at 1992Q4 the steady state prior for variable $i$ is row $i$ column $1$ + column $2$. Now lets input our priors and then append them to 'stan_data'. We also need to attach the dummy variable.


```{r, fig.width=8, fig.height=8}
priors <- priors(yt, p, lambda1, lambda2, fol_pm, Psi_0, vec_Psi_vars, dummy=dum_var)
stan_data <- c(stan_data, priors)
```

Like in Villani (2009), to incorporate that Sweden is a small economy and therefore not likely to affect the foreign economy, we restrict the upper right submatrix in each $A_i, i =1,\dots,k$, to the zero matrix. 

```{r, fig.width=8, fig.height=8}
n1 <- 3 #first 3 variables are foreign in yt
n2 <- 4 #the other 4 are domestic
n <- stan_data$k

tmp <- matrix(1, n*p, n)

for(i in 1:p){
  rows <- ((i-1)*n + n1 + 1) : (i*n)
  cols <- 1:n1
  tmp[rows, cols] <- 0
  zero_indices <- which(c(tmp) == 0)
}
diag(stan_data$Sigma_vec_beta[zero_indices, zero_indices]) <- 0.00001

#example for A_1, we restrict the elements with zero entries to have very very small prior variance
#and since prior mean is zero, the posterior will be essentially zero 
t(tmp)[,1:7]
```

Now we supply our forecast horizon $H$, and also the deterministic variables for the future periods and then we fit the model.

```{r, fig.width=8, fig.height=8}
H <- 8
X_pred <- cbind(rep(1, H), 0)
stan_fit <- estimate(stan_data, n_chains=4, iter=5000, warmup=2500, H=H, X_pred=X_pred)
```

Let us look at the posterior mean of $\beta$, $\Psi$ and $\Sigma_u$

```{r, fig.width=8, fig.height=8}
params <- c("beta", "Psi", "Sigma_u")
posterior_draws = rstan::extract(stan_fit)

for (param in params) {
  draws <- posterior_draws[[param]]
  mean_matrix <- round(apply(draws, c(2,3), mean), 2)
  cat("\n", paste0(param, "_posterior_mean (stan estimation)"), "\n")
  print(mean_matrix)
}
```

Lets plot the forecasts. Lets select a 95% prediction interval and the mean of the posterior as the point forecast. For the variables in annualized quarter on quarter growth rates, we transform the historical data and predictions to yearly growth rates.

```{r, fig.width=8, fig.height=8}
plot_forecast(stan_fit,
              yt,
              ci=0.95,
              fcst_type="mean", 
              growth_rate_idx=c(1,2,4,5),
              plot_idx=c(4,5,6),
              gibbs=FALSE)
```

Lets estimate the model with a Gibbs sampler instead. The priors and everything is the same, except for $\Sigma_u$ we now use the noninformative prior $\left|\Sigma_u \right|^{-(k+1)/2}$

```{r, fig.width=8, fig.height=8}
gibbs_fit <- estimate_gibbs(stan_data, iter=10000, warmup=5000, H=H, X_pred=X_pred)
```

We can check the posterior means
```{r, fig.width=8, fig.height=8}
for (i in seq_along(params)) {
  param <- params[i]
  cat("\n", paste0(param, "_posterior_mean (gibbs sampler)"), "\n", "\n")
  print(round(gibbs_fit[[i + 1]], 2))
}
```

Very similar to the stan estimation. Now lets plot the forecasts
```{r, fig.width=8, fig.height=8}
plot_forecast(gibbs_fit,
              yt,
              ci=0.95,
              fcst_type="mean", 
              growth_rate_idx=c(1,2,4,5),
              plot_idx=c(4,5,6),
              gibbs=TRUE)
```