---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SteadyStateBVAR

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/markjwbecker/steady_state_bvar/graph/badge.svg)](https://app.codecov.io/gh/markjwbecker/steady_state_bvar)
[![R-CMD-check](https://github.com/markjwbecker/SteadyStateBVAR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/markjwbecker/SteadyStateBVAR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

With this package, the user can estimate the steady-state BVAR(p) model of Mattias Villani (Villani, 2009). After estimation, the user can produce forecasts (unconditional and conditional) and impulse response functions (orthogonalized and generalized). The goal is to use modern Bayesian tools (Stan) to: i) estimate the model as specified in the original paper, and ii) extend the model in different ways. Back in the day, extensions of the model seemed to be limited by what Mattias Villani had time to derive.

See for example, Clark (2011): "*In a methodological sense, this paper extends the estimator of Villani (2009) to include stochastic volatility.*" Later on, he writes: "*(Special thanks are due to Mattias Villani for providing the formulas for posterior means and variances of* $\Pi$ *and* $\Psi$*, which generalize the constant-variance formulas of Villani 2009.)*"

For a more recent example of the model being at the mercy of Professor Villani, we can take a look at the [Technical Guide](https://github.com/european-central-bank/BEAR-toolbox/blob/master/tbx/doc/Technical%20guide.pdf) for the BEAR (Bayesian Estimation, Analysis and Regression) toolbox by the ECB. On page 122, Dieppe, Legrand, and van Roye (2018) write: "*Villani (2009) only provides derivation in the case of the normal-diffuse prior distribution, so the incoming analysis will be restricted to this case.*"

Times are different now, and with the help of Stan, we can basically do whatever we can imagine. To showcase this, in the last section we build on the steady-state BVAR model of Clark (2011) [which itself is an extension of the original model (Villani, 2009)], by letting the log volatilities follow correlated AR(1) processes instead of uncorrelated driftless random walks. And this is done without asking Professor Villani for any derivations. I simply wrote down the model equations, put some priors on the parameters, and Stan did the rest!

## Installation

You can install the development version of SteadyStateBVAR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("markjwbecker/SteadyStateBVAR")
```

## Introduction

The steady-state BVAR($p$) model (Villani, 2009) is

$$
y_t = \Psi d_t + \Pi_1(y_{t-1}-\Psi d_{t-1})+\dots+\Pi_p(y_{t-p}-\Psi d_{t-p})+u_t
$$

where $y_t$ is a $k$-dimensional vector of endogenous variables (time series) at time $t$, $d_t$ is a $q$-dimensional vector of deterministic (exogenous) variables at time $t$, and the (reduced-form) innovations are $u_t \sim N_k(0,\Sigma_u)$ with independence between time periods. Here $\Pi_\ell$ for $\ell=1,\dots,p$ is a $(k \times k)$ matrix, and $\Psi$ is a $(k \times q)$ matrix. Now


$$
\mathrm{E}(y_t)=\mu_t=\Psi d_t
$$

is the unconditional mean, or the **steady state**, of the process. Since long-horizon forecasts from stationary VARs converge to the unconditional mean (steady state), it is naturally very important from a forecasting perspective to obtain precise inference on $\Psi$.

Note that the current version of this package only allows for $d_t$ to contain either a constant, a constant and a dummy variable, or a constant and a time trend. We can stack the (transposed) $\Pi_i$ matrices in the $(kp \times k)$ matrix $\beta$

$$
\beta=
\begin{bmatrix}
\Pi'_1 \\ 
\vdots  \\
\Pi'_p
\end{bmatrix}
$$

We can then rewrite the model as a nonlinear regression (Karlsson, 2013)

$$
y_t' =d_t'\Psi' + \left[w_t'-q_t'(I_p \otimes \Psi') \right]\beta +u_t'
$$

where $w_t'=(y_{t-1}',\dots,y_{t-p}')$ is a $kp$-dimensional vector of lagged endogenous variables and $q_t'=(d_{t-1}',\dots,d_{t-p}')$ is a $qp$-dimensional vector of lagged deterministic (exogenous) variables, $I_p$ is the $(p \times p)$ identity matrix and $\otimes$ denotes the Kronecker product. This is how the likelihood is written in the Stan code. The goal is to estimate $\beta, \Psi$, and $\Sigma_u$, so priors are needed. First, prior independence between $\beta, \Psi$ and $\Sigma_u$ is assumed. Starting with $\beta$, we use the Minnesota prior


$$
\mathrm{vec}(\beta) \sim \mathrm{N}_{kpk} \left[\theta_\beta,\Omega_\beta\right]
$$

The prior means (the elements of $\theta_\beta$) are set to


$$
\begin{aligned}
\mathrm{E}\left(\Pi_{\ell}^{(i,j)}\right)&=
\begin{cases}
\kappa & \text{if } \ell = 1 \ \mathrm{and} \ i = j \\
0 & \mathrm{otherwise}
\end{cases}\\
\kappa&=
\begin{cases}
\kappa^{level} & \text{if } \mathrm{variable} \ i \ \mathrm{is in level} \\
\kappa^{\Delta} & \text{if }\mathrm{variable} \ i \ \mathrm{is in difference}
\end{cases}\\
\end{aligned}
$$


Here, the autoregressive coefficient $\Pi_{\ell}^{(i,j)}$ is element $\left(i,j\right)$ of $\Pi_{\ell}$ for $\ell=1,\dots,p$. As such, the Minnesota prior sets all prior means for the elements in $\beta$ to $0$, except for the elements that relate to the first own lags of the variables, which are set to $\kappa$. If variable $i$ is in level (e.g. nominal interest rate), then $\kappa=\kappa^{level}$, and typical choices for $\kappa^{level}$ are $1$ or $0.9$. Evaluating the equations at their prior means, equation $i$ becomes a random walk if $\kappa^{level}=1$ and a persistent stationary AR(1) process if $\kappa^{level}=0.9$. Since the steady state only exists if the process is stationary, $0.9$ is recommended for the steady-state BVAR. If variable $i$ is in difference (e.g. output growth), then $\kappa=\kappa^{\Delta}$, and the most common choice for $\kappa^{\Delta}$ is $0$, i.e. equation $i$ becomes (when evaluating it at its prior means) a random walk expressed in first differences. If a differenced variable still shows some degree of persistence (can be examined with an ACF plot), a suitable value for $\kappa^{\Delta}$ can be (for example) $0.6$ instead of $0$. Moving on to the prior variances, $\Omega_\beta$ is a diagonal matrix containing the prior variances for the elements in $\beta$. They are specified as


$$
\mathrm{Var}\left(\Pi_{\ell}^{(i,j)}\right)=
\begin{cases}
\left(\frac{\lambda_1}{\ell^{\lambda_3}}\right)^2 & \text{if } i = j \\
\left(\frac{\lambda_1 \lambda_2\sigma_i}{\ell^{\lambda_3}\sigma_j}\right)^2& \text{if } i \neq j
\end{cases}
$$


Here $\lambda_1$, $\lambda_2$, and $\lambda_3$ are scalar hyperparameters known as the overall tightness, the cross-equation tightness and the lag decay rate. Furthermore, $\sigma_i^2$ is the $(i,i)$:th element of $\Sigma_u$, which we do not know and therefore replace with an estimate. In this package, it is replaced by the least squares residual variance from a univariate autoregression for variable $i$ with $p$ lags (including the constant and dummy/trend variable if applicable). Moving on to $\Psi$, the prior we use is


$$
\mathrm{vec}(\Psi) \sim \mathrm{N}_{kq}\left[\theta_\Psi,\Omega_\Psi\right]
$$

This is really the core of the steady-state BVAR model. In $\theta_\Psi$, we specify our prior beliefs about the location of the steady state, and in $\Omega_\Psi$, which we assume to be a diagonal matrix, we specify our degree of certainty in those prior beliefs. Finally, the prior for $\Sigma_u$ is the usual noninformative Jeffreys prior


$$
p(\Sigma_u) \propto\left|\Sigma_u \right|^{-(k+1)/2}
$$

However, if the user wants, an inverse Wishart prior can be used instead


$$
\Sigma_u \sim \mathrm{IW}(V_0,m_0)
$$


where $V_0$ is the scale matrix and $m_0\geq k+2$ is the number of degrees of freedom. An uninformative prior can be (and is in this package) specified by setting $V_0=(m_0-k-1)\hat{\Sigma}_u$ where $\hat{\Sigma}_u$ is the least squares estimate from the VAR($p$) (including the constant and dummy/trend variable if applicable), and $m_0=k+2$.

This package also allows for stochastic volatility (Random Walk or AR1 specifications), where we let the covariance matrix of the innovations vary over time such that we have a time-varying covariance matrix $\Sigma_{u,t}$ (more details on this found in the respective vignettes).

##Example

To estimate the model in Section 4.1 of Villani (2009) and perform impulse response analysis and forecasting, simply run the following code:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(SteadyStateBVAR)
data("Villani2009")
yt <- Villani2009
yt <- ts(yt[1:102, ], start = start(yt), frequency = frequency(yt))

bvar_obj <- bvar(data = yt)

bp <- which(time(yt) == 1992.75)
dum_var <- c(rep(1,bp), rep(0,nrow(yt)-bp))

bvar_obj <- setup(bvar_obj,
                  p=4,
                  deterministic = "constant_and_dummy",
                  dummy = dum_var)

lambda_1 <- 0.2
lambda_2 <- 0.5
lambda_3 <- 1.0

#fol_pm = first own lag prior means
fol_pm=c(0,   #delta y_f
         0,   #pi_f
         0.9, #i_f
         0,   #delta y
         0,   #pi
         0.9, #i
         0.9  #q
         )

#psi_1 = Psi col 1
#psi_2 = Psi col 2

# default is 95% prior probability ('interval=0.95')
theta_Psi <- 
  c(
  ppi( 2.00,  3.00,  annualized_growthrate=TRUE)$mean,   #psi_1: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$mean,   #psi_1: pi_f
  ppi( 4.50,  5.50,  annualized_growthrate=FALSE)$mean,  #psi_1: i_f
  ppi( 2.00,  2.50,  annualized_growthrate=TRUE)$mean,   #psi_1: delta y
  ppi( 1.70,  2.30,  annualized_growthrate=TRUE)$mean,   #psi_1: pi
  ppi( 4.00,  4.50,  annualized_growthrate=FALSE)$mean,  #psi_1: i
  ppi( 3.85,  4.00,  annualized_growthrate=FALSE)$mean,  #psi_1: q
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$mean,   #psi_2: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$mean,   #psi_2: pi_f
  ppi( 1.50,  2.50,  annualized_growthrate=FALSE)$mean,  #psi_2: i_f
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$mean,   #psi_2: delta y
  ppi( 4.30,  5.70,  annualized_growthrate=TRUE)$mean,   #psi_2: pi
  ppi( 3.00,  5.50,  annualized_growthrate=FALSE)$mean,  #psi_2: i
  ppi(-0.50,  0.50,  annualized_growthrate=FALSE)$mean   #psi_2: q
  )

Omega_Psi <- 
  diag(
  c(
  ppi( 2.00,  3.00,  annualized_growthrate=TRUE)$var,    #psi_1: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$var,    #psi_1: pi_f
  ppi( 4.50,  5.50,  annualized_growthrate=FALSE)$var,   #psi_1: i_f
  ppi( 2.00,  2.50,  annualized_growthrate=TRUE)$var,    #psi_1: delta y
  ppi( 1.70,  2.30,  annualized_growthrate=TRUE)$var,    #psi_1: pi
  ppi( 4.00,  4.50,  annualized_growthrate=FALSE)$var,   #psi_1: i
  ppi( 3.85,  4.00,  annualized_growthrate=FALSE)$var,   #psi_1: q
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$var,    #psi_2: delta y_f
  ppi( 1.50,  2.50,  annualized_growthrate=TRUE)$var,    #psi_2: pi_f
  ppi( 1.50,  2.50,  annualized_growthrate=FALSE)$var,   #psi_2: i_f
  ppi(-1.00,  1.00,  annualized_growthrate=TRUE)$var,    #psi_2: delta y
  ppi( 4.30,  5.70,  annualized_growthrate=TRUE)$var,    #psi_2: pi
  ppi( 3.00,  5.50,  annualized_growthrate=FALSE)$var,   #psi_2: i
  ppi(-0.50,  0.50,  annualized_growthrate=FALSE)$var    #psi_2: q
  )
  )

bvar_obj <- priors(bvar_obj,
                   lambda_1,
                   lambda_2,
                   lambda_3,
                   fol_pm,
                   theta_Psi,
                   Omega_Psi,
                   Jeffrey=TRUE)

p <- bvar_obj$setup$p
k <- bvar_obj$setup$k
kf <- 3 #foreign variables

restriction_matrix <- matrix(1, k*p, k)

for(i in 1:p){
  rows <- ((i-1)*k + kf + 1) : (i*k)
  cols <- 1:kf
  restriction_matrix[rows, cols] <- 0
}
#block exogeneity for foreign variables
bvar_obj <- restrict_beta(bvar_obj, restriction_matrix)

bvar_obj$predict$H <- 12
bvar_obj$predict$d_pred <- cbind(rep(1, 12), 0)

bvar_obj <- fit(bvar_obj,
                iter = 20000,
                warmup = 10000,
                chains = 4)

summary(bvar_obj)

fcst <- forecast(bvar_obj,
                 ci = 0.95,
                 fcst_type = "mean",
                 growth_rate_idx = c(4,5), #annual inflation/real GDP growth instead of QoQ
                 plot_idx = c(4,5,6))

irf <- IRF(bvar_obj,
           H=20,
           response=5,#inflation
           shock=6, #interest rate
           type="median",
           method="OIRF",
           ci=0.68,
           growth_rate_idx=5) #annual inflation instead of QoQ
```

## References

Clark, T. E. (2011). Real-Time Density Forecasts from Bayesian Vector Autoregressions
with Stochastic Volatility. *Journal of Business \& Economic Statistics*. 29(3), pp. 327–341.

Dieppe, A., Legrand, R., and van Roye, B. (2018). *The Bayesian Estimation, Analysis and Regression (BEAR) Toolbox Technical guide*. European Central Bank.

Karlsson, S. (2013). Forecasting with Bayesian Vector Autoregression. In: Elliott, G. and Timmerman, A. (eds) *Handbook of Economic Forecasting*. Elsevier B.V. Vol 2, Part B., pp. 791-897.

Villani, M. (2009). Steady-state priors for vector autoregressions. *Journal of Applied Econometrics*. 24(4), pp. 630-650. 
